---
title: "hw2-exploring-patterns-of-envi-injust"
format: html
---

```{r}
# Load needed libraries 
library(tidyverse)
library(sf) # For handling vector data
library(tmap) # For making maps
library(spData) # Pre-loaded spatial data
```

Read in data
```{r}
# Read in environmental justice data
ejscreen <- st_read(here::here("data", "ejscreen", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))

# Read in redlining data
redlining <- st_read(here::here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))

# Read in biodiversity data
bio_div <- st_read(here::here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))
```
Select for Los Angeles County data: 
```{r}
 la_ejscreen <- ejscreen |> 
  filter(CNTY_NAME == "Los Angeles County")
```

Looking at the CRS for all of the datasets: 
```{r}
# EJ screen CRS
st_crs(la_ejscreen)
```


```{r}
# Redlining CRS
st_crs(redlining)
```


```{r}
# Biodiversity CRS
st_crs(bio_div)
```
```{r}

```
 Check to see if the projection is geographic or projected.
```{r}
st_crs(la_ejscreen)$IsGeographic
st_crs(redlining)$IsGeographic
st_crs(bio_div)$IsGeographic
```
 
Since the environmental justice data is projected, and projected coordinate reference systems are better for 2D map making this is the CRS we will keep. Change the CRS of redlining data and biodiversity data to match the CRS of the environmental justice screen data.  
```{r}
# Use st_transform() to set the CRS to that of the environmental justice data
redlining <- st_transform(redlining, crs = st_crs(la_ejscreen))
bio_div <- st_transform(bio_div, crs = st_crs(la_ejscreen))
```

Making sure the CRS are the same for all of the datasets
```{r}
# Setting the CRS to equal each other in a boolean to make sure they are all the same
# Check relining data is using the same CRS as la_ejscreen data
if (st_crs(la_ejscreen)==st_crs(redlining)){
  print("it's a match!")
}else {
    print("still not a match")
}

# Check bio diversity data is using the same CRS as la_ejscreen data
if (st_crs(la_ejscreen)==st_crs(bio_div)){
  print("it's a match!")
}else {
    print("still not a match")
  }
```


# Part 1: Legacy of Redlining in current environmental (in)justice
1. Create a map if historical relining neighborhoods 
```{r}
tm_shape(la_ejscreen) +
  tm_polygons() + 
tm_shape(redlining, is.main = TRUE) + # is.main to make this the size of the map
  tm_polygons()
```

